*Prochlorococcus marinus* responses to light and oxygen

### 2024-03-01 Code/Manuscript_O2.Rmd
minor edits

Three xenic cultures of *P. marinus* were obtained from Bigelow Labs, NCMA
Maine, US (REFERENCE). MED4 (CCMP1986) is from High-Light adapted (HLI) clade; SS120
(CCMP1375) is from Low-Light adapted (LLIII) clade; and MIT9313 (CCMP2773)
is from Low-Light adapted (LLIV) clade. Cultures were
maintained in incubators set to 22°C with a light/dark cycle of 12 h. The PAR level for maintenance cultures reflected PAR)in the source niche of the ecotype; MED4, of 160 µmol photons m^-2^ s^-1^ with illumination from STANDARD Products Inc. Cool White F24T5/41K/8/HO/PS/G5/STD, 24 watts, fluorescent bulbs; SS120 and
MIT9313 at 30 µmol photons m^-2^ s^-1^ with illumination from Philips Cool White F14T5/841 Alto, 14 watts, fluorescent bulbs. To maintain active growth all strains were
transferred weekly with 1 in 5 dilutions with Pro99 media
[@mooreCulturingMarineCyanobacterium2007] prepared with autoclaved artificial
seawater [@bergesEvolutionArtificialSeawater2001].

@@

## Growth rate analysis {.unnumbered}
Data files (.csv) saved from the Multicultivator software were imported into R-Studio for data management [@R-tidyverse], growth rate calculations CITATION, comparisons of model fits [@harrisonPhotosynthesisirradianceRelationshipsPolar1986], and visualization. The chlorophyll proxy
optical density (OD~680~ - OD~720~;  ΔOD) was used to determine the chlorophyll specific growth
rate (µ, d^-1^) for each treatment combination.  We first used a rolling mean from the R package *`zoo`* [@R-zoo] to calculate the average ΔOD data over a 1-hour window to lower the influence of outlier points and remove data points collected during post stationary phase, when applicable. We used the Levenberg-Marquardt algorithm
[@bellaviaLevenbergMarquardtMethod2018] modification of the non-linear
least squares,  using the R package *`minpack.lm`*
[@R-minpack.lm], to fit a logistic
equation (Equation \@ref(eq:GrowthRate))

@@

where ΔOD~max~ is maximum ΔOD, ΔOD~min~ is minimum ΔOD, t is time
duration over the growth trajectory.

@@

An integrated weighting equation \@ref(eq:WeightedPUR) [@morelAvailableUsableStored1978] was used to determine the weighted PUR spectrum P(λ),

@@

where A(λ) is the blank subtracted, Chl a~2~ peak normalized whole cell absorbance spectrum for each *P. marinus* ecotype, over 400 nm to 700 nm, A(λ); and E(λ) is the peak normalized emission spectrum of the imposed LED growth light, over 400 nm to 700 nm.

Actinic PUR levels (µmol photons m^-2^ s^-1^) were calculated from actinic PAR (µmol photons m^-2^ s^-1^) levels using the equation \@ref(eq:PUR) from [@morelAvailableUsableStored1978],

@@

where P(λ) is the weighted PUR absorbance spectrum from equation \@ref(eq:WeightedPUR), E(λ) is the imposed growth light emission spectrum from equation \@ref(eq:WeightedPUR) and PAR is the imposed peak light level (µmol photons m^-2^ s^-1^).  Figure (\@ref(fig:PurParPlot)) shows the calculated absorbed peak PUR (µmol photons m^-2^ d^-1^) versus imposed actinic peak PAR (µmol photons m^-2^ s^-1^) for each strain and each spectral waveband (nm).

The applied photoperiods were delivered using the sine circadian light function of the PSI Multicultivator to simulate light exposure approximating sun rise through to sunset. The area under the sinusoidal curves is equivalent to the area of a triangular photoregime of equivalent photoperiod (Campbell, unpub), therefore the equation to determine the cumulative diel PUR (µmol photons m^-2^ d^-1^) is one half of the base (photoperiod) multiplied by the height (PUR) (Equation \@ref(eq:SumPURDay)),

@@

where PUR is the actinic absorbed light calculated from equation \@ref(eq:PUR) (µmol photons m^-2^ s^-1^), 3600 is the time conversion from seconds to hour and photoperiod is the imposed photoperiod (h).

@@

We fit the response of chlorophyll specific growth rate to cumulative diel PUR using the equation of Harrison and Platt [@harrisonPhotosynthesisirradianceRelationshipsPolar1986] across strain and [O~2~] to first compare growth response between 660 nm (red) and 450 nm (blue) growth light and second to compare growth response between specific photoperiods (4 h, 8 h, 12 h, 16 h) and a fit across the pooled photoperiod data. 
To examine statistical differences between the modeled fits, we performed one-way ANOVA comparing the model output parameters assigning significant differences when p < 0.05.

@@

Guided by the evidence of ocean distributions of proteins from *Prochlorococcus* we set up a matrix of photoperiods, PAR, spectral bands, and [O~2~] to approximate current, and potential future, latitudinal, depth and seasonal niches for *Prochlorococcus* strains.  We earlier found [@murphyPhotoinactivationPhotosystemII2017] that growth under red light could prove mechanistically informative as to factors limiting *Prochlorococcus* growth, so we included that spectral band even though it is not characteristic of representative *Prochlorococcus* niches.  Although *Prochlorococcus* is currently limited to a narrow range of surface photoperiods, potential poleward latitudinal expansions, in combination with attenuation of light with depth, mean *Prochlorococcus* may potentially encounter a wide range of photoperiods.  Our growth rate determinations generally agreed with those from Moore *et al*. [@mooreComparativePhysiologySynechococcus1995], for white LED and 250 µM O~2~, but our study is, to our knowledge, the first to analyze the interactive growth responses of *Prochlorococcus* strains to varying [O~2~], spectral wavebands and photoperiods.

@@

Figure \@ref(fig:ProchlorococcusEnzymeKms) shows genes encoding enzymes for  *P. marinus* strains from Clades HLI, LLI, LLIII and LLIV [@omarAnnotationGenesEncoding2023], compared to the measured or inferred K~M~ for [O~2~] for each enzyme CITATIONS. MED4 increases expression of alternative oxidase ('ubiquinol oxidase (non electrogenic)')  to cope with changes in light [@bergResponsesPsbAHli2011], by dissipating electrons from the inter-system transport chain. The approximate K~M~ for [O~2~] of ~ 25 µM for ubiquinol oxidase (non electrogenic) is comparable to the lower limit for growth of MED4 in our experiments.  We suggest that dependence upon this enzyme excludes MED4 from low oxygen zones.  The genome scan shows SS120 and MIT9313 lack this gene, and therefore, lack this oxygen-dependent path to cope with changing excitation.  Conversely, a gene encoding (S)-2-hydroxy-acid oxidase is encoded in the MIT9313 genome.  (S)-2-hydroxy-acid oxidase catalyzes the reaction of 2-hydroxy acid with O~2~ to produce toxic H~2~O~2~ [@cunaneCrystalStructureAnalysis2005]. (S)-2-hydroxy-acid oxidase has an approximate K~M~ for [O~2~] of ~ 250 µM, and produces H~2~O~2~, so growth at lower [O~2~] may protect MIT9313 from auto-intoxication from production of H~2~O~2~. We hypothesize that under 250 µM O~2~ and higher blue light, *P. marinus* MIT9313 suffered photoinhibition, resulting from the inactivation of PSII caused by the production of the reactive oxygen species, hydrogen peroxide. This photoinhibition is compounded by the limited inducible repair mechanism for PSII, due to the absence of FtsH 2 and 3  expression in *P. marinus* MIT9313 [@bonisteelStrainSpecificDifferences2018].  We hypothesize that under the conditions of our high light and 2.5 µM and 25 µM O~2~ experiments, the activity of the (S)-2-hydroxy-acid oxidase enzyme is suppressed. As a result, the catalyzed production of hydrogen peroxide is inhibited, leading to less PSII damage, allowing MIT9313 to avoid photoinhibition and circumvent its limitations on PSII repair.  Figure \@ref(fig:ProchlorococcusEnzymeKms) also shows that *P. marinus* SS120 is the only tested ecotype to lack the pyridoxal 5'-phosphate synthase enzyme. The pyridoxal 5'-phosphate synthase enzyme is an important cofactor in the biosynthesis of vitamin B~6~ [@sangIdentificationPyridoxinePyridoxamine2007]. Vitamin B~6~ is a potential antioxidant and can effectively quench singlet oxygen [@bilskiVitaminB6Pyridoxine2000]. The absence of the pyridoxal 5'-phosphate synthase enzyme may explain why *P. marinus* SS120 does not grow as well as *P. marinus* MIT9313, when exposed to high light stress under 25 µM O~2~ and not at all under 2.5 µM O~2~.

@@

Figure \@ref(fig:DNARepairFig) shows genes encoding DNA repair for *P. marinus* strains. As expected, *P. marinus* MED4 possesses the largest, most complete suite of genes encoding DNA repair enzymes, followed by *P. marinus* MIT9313. Conversely, *P. marinus* SS120 demonstrates the smallest genomic capacity for DNA repair. *Prochlorococcus marinus* MED4 was the only strain to possess a gene encoding deoxyribodipyrimidine photolyase, which, in the presence of blue light, is responsible for repairing DNA damaged by UV light [@sancarStructureFunctionDNA2003]. *Prochlorococcus marinus* MED4 was also the only strain to possesses a gene encoding DNA ligase, which uses ATP as a cofactor for DNA repair. The absence of  genes encoding deoxyribodipyrimidine photolyase and DNA ligase (ATP) in *P. marinus* MIT9313 and *P. marinus* SS120 helps explain why these two strains cannot tolerate growth under full [O~2~] and high light, found at the ocean surface. Furthermore, the protective effect of lower [O~2~], allowing these strains to grow at higher light, may relate in part to suppression of DNA damage when generation of Reactive Oxygen Species is suppressed at lower [O~2~].  *Prochlorococcus* are highly susceptible to hydrogen peroxide (H~2~O~2~) toxicity as they lack genes which scavenge H~2~O~2~ molecules [@morrisFacilitationRobustGrowth2008]. The small cell size of *Prochlorococcus* allow the reactive oxygen species (ROS), H~2~O~2~, to cross the cell membrane [@omarGenomicCapacitiesReactive2023]; however, accumulation of extracellular H~2~O~2~ remains toxic to *Prochlorococcus* [@morrisFacilitationRobustGrowth2008,  @morrisDependenceCyanobacteriumProchlorococcus2011].

@@

*Prochlorococcus marinus* MED4 has a physiological requirement for more than 4 h of light per day, and thus this strain will not exploit habitats with short photoperiods, typical of winter or light attenuated depths. MED4 is also excluded from the lowest oxygen habitats, equivalent to our 2.5 µM O~2~ experimental conditions. MED4 can, however, grow under OMZ regions with slightly higher [O~2~], as demonstrated by the growth under our 25 µM O~2~ experiments. Bioinformatic analyses  (Figure \@ref(fig:ProchlorococcusEnzymeKms))  and previous transcriptional analyses [@bergResponsesPsbAHli2011]  suggest MED4 is excluded from growth below ~ 25 µM O~2~ because it relies upon a ubiquinol oxidase, non-electrogenic, to maintain oxidation/reduction balance in the intersystem electron transport chain.  On the other hand, MED4 shows inducible expression of FtsH isoforms [@bonisteelStrainSpecificDifferences2018], to counter photoinactivation of PSII under higher PAR and [O~2~] environments. Photoinactivation, does, however, impose an increased cost of growth upon MED4, since growth under red light, to lower the rate of photoinactivation of PSII [@murphyPhotoinactivationPhotosystemII2017], allows MED4 to achieve faster growth per absorbed photon than growth under red light. TARA Oceans
Project data [@delmontLinkingPangenomesMetagenomes2018] indeed reported presence of *P. marinus* MED4-like genomes at depths ranging from 5 m to 90 m, representing high to low blue light levels, in the Pacific South East Ocean. @delmontLinkingPangenomesMetagenomes2018 did not analyze data from depths beyond the subsurface chlorophyll maximum layer, the layer in the water column where the chlorophyll *a* concentration peaks, nor did they report the [O~2~] at those depths. Our growth findings are also consistent with Figure\@ref(fig:NATLProchloroPhotosynthDepthO2) showing PSII proteins annotated as MED4, at depths up to 200 meters with O~2~ of 15 µM.

>>>>

Three xenic cultures of *P. marinus* were obtained from Bigelow Labs, NCMA
Maine, USA. MED4 (CCMP1986) is from High-Light adapted (HLI) clade; SS120
(CCMP1375) is from Low-Light adapted (LLIII) clade; and MIT9313 (CCMP2773)
is from Low-Light adapted (LLIV) clade. Cultures were
maintained in incubators set to 22°C with a light/dark cycle of 12 h. The PAR level for maintenance cultures reflected PAR)in the source niche of the ecotype; MED4, of 160 µmol photons m^-2^ s^-1^ with illumination from STANDARD Products Inc. Cool White F24T5/41K/8/HO/PS/G5/STD, 24 watts, fluorescent bulbs; SS120 and
MIT9313 at 30 µmol photons m^-2^ s^-1^ with illumination from Philips Cool White F14T5/841 Alto, 14 watts, fluorescent bulbs. To maintain active growth all strains were
transferred weekly with 1 in 5 dilutions with Pro99 media
[@mooreCulturingMarineCyanobacterium2007] prepared with autoclaved artificial
seawater [@bergesEvolutionArtificialSeawater2001].

@@

## Growth rate analysis {.unnumbered}
Data files (.csv) saved from the Multicultivator software were imported into R-Studio for data management [@R-tidyverse], growth rate calculations CITATION, comparisons of model fits [@harrisonPhotosynthesisirradianceRelationshipsPolar1986], and visualization. The chlorophyll proxy
optical density (OD~680~ - OD~720~;  ΔOD) was used to determine the chlorophyll specific growth
rate (µ, d^-1^) for each treatment combination.  We first used a rolling mean from the R package *`zoo`* [@R-zoo] to calculate the average ΔOD data over a 1-hour window to lower the influence of outlier points and remove data points collected during post stationary phase, when applicable. We used the Levenberg-Marquardt algorithm
[@bellaviaLevenbergMarquardtMethod2018] modification of the non-linear
least squares,  using the R package *`minpack.lm`*
[@R-minpack.lm], to fit a logistic
equation (Equation \@ref(eq:GrowthRate)); where ΔOD~max~ is maximum ΔOD, ΔOD~min~ is minimum ΔOD, t is time
duration over the growth trajectory.

@@

@@

An integrated weighting equation \@ref(eq:WeightedPUR) [@morelAvailableUsableStored1978] was used to determine the weighted PUR spectrum P(λ); where A(λ) is the blank subtracted, Chl a~2~ peak normalized whole cell absorbance spectrum for each *P. marinus* ecotype, over 400 nm to 700 nm, A(λ); and E(λ) is the peak normalized emission spectrum of the imposed LED growth light, over 400 nm to 700 nm.

@@

Actinic PUR levels (µmol photons m^-2^ s^-1^) were calculated from actinic PAR (µmol photons m^-2^ s^-1^) levels using the equation \@ref(eq:PUR) from [@morelAvailableUsableStored1978]; where P(λ) is the weighted PUR absorbance spectrum from equation \@ref(eq:WeightedPUR), E(λ) is the imposed growth light emission spectrum from equation \@ref(eq:WeightedPUR) and PAR is the imposed peak light level (µmol photons m^-2^ s^-1^).  Figure (\@ref(fig:PurParPlot)) shows the calculated absorbed peak PUR (µmol photons m^-2^ d^-1^) versus imposed actinic peak PAR (µmol photons m^-2^ s^-1^) for each strain and each spectral waveband (nm).

@@

The applied photoperiods were delivered using the sine circadian light function of the PSI Multicultivator to simulate light exposure approximating sun rise through to sunset. The area under the sinusoidal curves is equivalent to the area of a triangular photoregime of equivalent photoperiod (Campbell, unpub), therefore the equation to determine the cumulative diel PUR (µmol photons m^-2^ d^-1^) is one half of the base (photoperiod) multiplied by the height (PUR) (Equation \@ref(eq:SumPURDay)); where PUR is the actinic absorbed light calculated from equation \@ref(eq:PUR) (µmol photons m^-2^ s^-1^), 3600 is the time conversion from seconds to hour and photoperiod is the imposed photoperiod (h).

@@

@@

We fit the response of chlorophyll specific growth rate to cumulative diel PUR using the equation of Harrison and Platt [@harrisonPhotosynthesisirradianceRelationshipsPolar1986] across strain and [O~2~] to compare growth response between 660 nm (red) and 450 nm (blue) growth light and additionally, to compare growth response between specific photoperiods (4 h, 8 h, 12 h, 16 h) and a fit across pooled photoperiod data. 
To examine statistical differences between the modeled fits, we performed one-way ANOVA comparing the model output parameters assigning significant differences when the p value was < 0.05.

@@

Guided by the evidence of ocean distributions of proteins from *Prochlorococcus* we set up a matrix of photoperiods, PAR, spectral bands, and [O~2~] to approximate current, and potential future, latitudinal, depth and seasonal niches for *Prochlorococcus* strains.  As aforementioned, growth under red light could prove mechanistically informative [@murphyPhotoinactivationPhotosystemII2017] to factors limiting *Prochlorococcus* growth, we therefore included the red spectral waveband even though it is not characteristic of representative *Prochlorococcus* niches.  Although *Prochlorococcus* is currently limited to a narrow range of surface photoperiods, potential poleward latitudinal expansions, in combination with attenuation of light with depth, mean *Prochlorococcus* may potentially encounter a wide range of photoperiods.  Our growth rate determinations generally agreed with those from Moore *et al*. [@mooreComparativePhysiologySynechococcus1995], for white LED and 250 µM O~2~, but our study is, to our knowledge, the first to analyze the interactive growth responses of *Prochlorococcus* strains to varying [O~2~], spectral wavebands and photoperiods.

@@

Figure \@ref(fig:ProchlorococcusEnzymeKms) shows genes encoding enzymes for  *P. marinus* strains from clades HLI, LLI, LLIII and LLIV [@omarAnnotationGenesEncoding2023], compared to the measured or inferred K~M~ for [O~2~] for each enzyme XXXX CITATION  to Table once added XXXX. MED4 increases expression of alternative oxidase ('ubiquinol oxidase (non electrogenic)')  to cope with changes in light [@bergResponsesPsbAHli2011], by dissipating electrons from the inter-system transport chain. The approximate K~M~ for [O~2~] of ~ 25 µM for ubiquinol oxidase (non electrogenic) is comparable to the lower limit for growth of MED4 in our experiments.  We suggest that dependence upon this enzyme excludes MED4 from low oxygen zones.  The genome scan shows SS120 and MIT9313 lack this gene, and therefore, lack this oxygen-dependent path to cope with changing excitation.  Conversely, a gene encoding (S)-2-hydroxy-acid oxidase is encoded in the MIT9313 genome.  (S)-2-hydroxy-acid oxidase catalyzes the reaction of 2-hydroxy acid with O~2~ to produce toxic H~2~O~2~ [@cunaneCrystalStructureAnalysis2005]. (S)-2-hydroxy-acid oxidase has an approximate K~M~ for [O~2~] of ~ 250 µM, and produces H~2~O~2~, so growth at lower [O~2~] may protect MIT9313 from auto-intoxication from production of H~2~O~2~. We hypothesize that under 250 µM O~2~ and higher blue light, *P. marinus* MIT9313 suffered photoinhibition, resulting from the inactivation of PSII caused by the production of the reactive oxygen species, hydrogen peroxide. This photoinhibition is compounded by the limited inducible repair mechanism for PSII, due to the absence of FtsH 2 and 3  expression in *P. marinus* MIT9313 [@bonisteelStrainSpecificDifferences2018].  We hypothesize that under the conditions of our high light and 2.5 µM and 25 µM O~2~ experiments, the activity of the (S)-2-hydroxy-acid oxidase enzyme is suppressed. As a result, the catalyzed production of hydrogen peroxide is inhibited, leading to less PSII damage, allowing MIT9313 to avoid photoinhibition and circumvent its limitations on PSII repair.  Figure \@ref(fig:ProchlorococcusEnzymeKms) also shows that *P. marinus* SS120 is the only tested ecotype to lack the pyridoxal 5'-phosphate synthase enzyme. The pyridoxal 5'-phosphate synthase enzyme is an important cofactor in the biosynthesis of vitamin B~6~ [@sangIdentificationPyridoxinePyridoxamine2007]. Vitamin B~6~ is a potential antioxidant and can effectively quench singlet oxygen [@bilskiVitaminB6Pyridoxine2000]. The absence of the pyridoxal 5'-phosphate synthase enzyme may explain why *P. marinus* SS120 does not grow as well as *P. marinus* MIT9313, when exposed to high light stress under 25 µM O~2~ and not at all under 2.5 µM O~2~.

@@

Figure \@ref(fig:DNARepairFig) shows genes encoding DNA repair for *P. marinus* strains. As expected, *P. marinus* MED4 possesses the largest, most complete suite of genes encoding DNA repair enzymes, followed by *P. marinus* MIT9313. Conversely, *P. marinus* SS120 demonstrates the smallest genomic capacity for DNA repair. *Prochlorococcus marinus* MED4 was the only strain to possess a gene encoding deoxyribodipyrimidine photolyase, which, in the presence of blue light, is responsible for repairing DNA damaged by UV light [@sancarStructureFunctionDNA2003]. *Prochlorococcus marinus* MED4 was also the only strain to possesses a gene encoding DNA ligase, which uses ATP as a cofactor for DNA repair. The absence of  genes encoding deoxyribodipyrimidine photolyase and DNA ligase (ATP) in *P. marinus* MIT9313 and *P. marinus* SS120 helps explain why these two strains cannot tolerate growth under full [O~2~] and high light, found at the ocean surface. Furthermore, the protective effect of lower [O~2~], allowing these strains to grow at higher light, may relate in part to suppression of DNA damage when generation of Reactive Oxygen Species is suppressed at lower [O~2~].  *Prochlorococcus* are highly susceptible to hydrogen peroxide (H~2~O~2~) toxicity as they lack genes which scavenge H~2~O~2~ molecules [@morrisFacilitationRobustGrowth2008]. The small cell size of *Prochlorococcus* allow the reactive oxygen species (ROS), H~2~O~2~, to cross the cell membrane [@omarDiffusionalInteractionsMarine2022]; however, accumulation of extracellular H~2~O~2~ remains toxic to *Prochlorococcus* [@morrisFacilitationRobustGrowth2008,  @morrisDependenceCyanobacteriumProchlorococcus2011].

@@

*Prochlorococcus marinus* MED4 has a physiological requirement for more than 4 h of light per day, and thus this strain will not exploit habitats with short photoperiods, typical of winter or light attenuated depths. MED4 is also excluded from the lowest oxygen habitats, equivalent to our 2.5 µM O~2~ experimental conditions. MED4 can, however, grow under OMZ regions with slightly higher [O~2~], as demonstrated by the growth under our 25 µM O~2~ experiments. Genomic analyses (Figure \@ref(fig:ProchlorococcusEnzymeKms))  and previous transcriptional analyses [@bergResponsesPsbAHli2011]  suggest MED4 is excluded from growth below ~ 25 µM O~2~ because it relies upon a ubiquinol oxidase, non-electrogenic, to maintain oxidation/reduction balance in the intersystem electron transport chain.  On the other hand, MED4 shows inducible expression of FtsH isoforms [@bonisteelStrainSpecificDifferences2018], to counter photoinactivation of PSII under higher PAR and [O~2~] environments. Photoinactivation, does, however, impose an increased cost of growth upon MED4, since growth under red light, to lower the rate of photoinactivation of PSII [@murphyPhotoinactivationPhotosystemII2017], allows MED4 to achieve faster growth per absorbed photon than growth under red light. TARA Oceans
Project data [@delmontLinkingPangenomesMetagenomes2018] indeed reported presence of *P. marinus* MED4-like genomes at depths ranging from 5 m to 90 m, representing high to low blue light levels, in the Pacific South East Ocean. @delmontLinkingPangenomesMetagenomes2018 did not analyze data from depths beyond the subsurface chlorophyll maximum layer, the layer in the water column where the chlorophyll *a* concentration peaks, nor did they report the [O~2~] at those depths. Our growth findings are also consistent with Figure\@ref(fig:NATLProchloroPhotosynthDepthO2) showing PSII proteins annotated as MED4, at depths up to 200 meters with O~2~ of 15 µM.
